Nail: A practical interface generator for data formats

### 2014-04-02 paper/design.tex
pass over design up to 3.2

\paragraph{Overview}

To alleviate these problems, Nail provides a richer set of combinators that not
only describes the grammar of the external protocol, but also induce an internal
model.

A central design decision of Nail is that there is a semantic bijection between
the model exposed to the programmer and the bytes input and output up to
syntactic equivalence for unambiguous grammars. More precisely, the parser is
the generators inverse, so parsing the generators output will yield the
generators input, but generating the parsers output does not necessarily yield
the parsers input. To understand why this makes sense, consider a grammar for a
text language that tolerates white space or a binary protocol that tolerates
arbitrarily long padding\footnote{Say, the physical layer of most communication
  protocols is a possibly infinite sequence of symbols that are syntactically
  nil followed by a pre-determined synchronisation sequence and the actual
  contents of the transmission.}. In that case, the program semantics should be
independent of the number of padding elements in the input, and Nail therefore
does not expose that information to the programmer. We call such discarded
fields \emph{constants}. Currently, Nail always makes a default choice when
there are multiple options to express a constant, however Nail could be extended
to allow a grammar-specific plug-in to make these choices, say for faster
alignment, streaming applications when data is not ready or visual appearance in
human-readable protocols. Similarly, Nail does not preserve the layout of
objects referred to by their offsets. If the grammar contains no constants and
offset fields, there is a proper isomorphism between model and protocol.

Nail evaluates the combinators and produces type declarations for the internal
model, the \textit{parser}, a function to parse a sequence of bytes into an
instance of the model and the \textit{generator}, a function to create a
sequence of bytes from an instance of the model. In the following, we present
Nail's combinators.

\paragraph{Fundamental parsers}
The elementary parsers of Nail are the same as those of Hammer, signed and
unsigned integers with arbitrary lengths up to 64 bits. Note that is possible to
define parsers for sub-byte lengths, e.g. to parse the 4-bit data offset within
the TCP header. Integer parsers return their value in the
smallest appropriately sized machine integer type, e.g. a 24 bit integer is stored in a 32-bit wide variable.\\
Examples: \texttt{uint4}, \texttt{int32}

Integer parsers can be constrained to fall either within an (inclusive) range of
values or be an
element of a set of allowed values. \\
Examples:\\
\texttt{uint8 | 1..3} (8 bit unsigned integer $1\leq x \leq 3$) ,\\
\texttt{uint16| ..512} (unsigned 16-bit integer $\leq 512$),\\
\texttt{int32 | [1,255,512]} (signed 32 bit integer with value 1, 255 or 512).
Invalid values or prematurely reaching the end of input raise an error when
parsing input or generating output and abort the procedure.

\paragraph{Sequence: Structure}
Nails fundamental concept is the structure combinator. It contains a list of
named parsers and unnamed constant parsers  . The parser invokes each field 
in sequence and returns a structure containing the result of each parser.

\begin{tabular}{ll}
  \textbf{Nail} & \textbf{Data Model}\\
  \begin{minipage}{1.2in}
\begin{verbatim}
header = {
 id uint16
 qr uint1
 opcode uint4
 aa uint1 
 tc uint1
 rd uint1
 ra uint1
 uint3 = 0
 rcode uint4
}
\end{verbatim} 
  \end{minipage}
  & 
  \begin{minipage}{1.2in}
\begin{verbatim}
struct header{
    uint16_t id;
    uint8_t qr;
    uint8_t opcode;
    uint8_t aa;
    uint8_t tc;
    uint8_t rd;
    uint8_t ra;
    uint8_t rcode;
};
\end{verbatim} 
  \end{minipage}
  \\
\end{tabular}
\paragraph{Repetition: Many,SepBy}
The many combinator takes a parser and applies it repeatedly until it fails,
returning an array of the inner parsers results. SepBy additionally takes a
constant parser, which it applies in between
parsing two values,but not before parsing the first value or after parsing the last.\\
Examples: \texttt{many uint8}\\ \texttt{sepBy uint8=',' (many uint8| '0'..'9')}
(which recognises comma-separated lists of numbers).

\paragraph{Semantic choice: choose}
We extend PEG's ordered choice combinator with a tag for each choice. The parser
attempts to parse each option in the order they are specified in the field and
stores the result in a tagged union. If an option fails, the parser backtracks
to the beginning of the choice combinator's input. Care must be taken that the
choices do not overlap, because Nail always picks to the first successfully
parsed choice. If two options overlap, generated output for the latter option is
not necessarily understood identically by the parser. However, actual data
formats are usually not ambiguous in this sense.

\begin{verbatim}
choose {
   A = uint8 | 1.. 8
   B = uint16 | ..256
}
\end{verbatim}

\subsection{Constant parsers}
Nail also features \emph{constant parsers}, which are not affecting the internal
representation. Constant parsers can appear instead of structure fields and as
separators in the SepBy combinator.

>>>>

\subsection{Overview}

To integrate the data format and the internal representation, Nail
provides a rich set of combinators that not
only describe the grammar of the external protocol, but also induce an internal
model.

A central design decision of Nail is that there is a \emph{semantic bijection} between
the model exposed to the programmer and the byte-level input and output, up to
syntactic equivalence for unambiguous grammars. More precisely, the parser is
the generator's inverse, so parsing the generator's output will yield the
generator's input, but generating the parsers output does not necessarily yield
the parsers input. To understand why this makes sense, consider a grammar for a
text language that tolerates white space, or a binary protocol that tolerates
arbitrarily long padding.\footnote{Say, the physical layer of most communication
  protocols is a possibly infinite sequence of symbols that are syntactically
  nil followed by a pre-determined synchronisation sequence and the actual
  contents of the transmission.} In that case, the program semantics should be
independent of the number of padding elements in the input, and Nail therefore
does not expose that information to the programmer. We call such discarded
fields \emph{constants}. Currently, Nail always makes a default choice when
there are multiple options to express a constant, however Nail could be extended
to allow a grammar-specific plug-in to make these choices, say for faster
alignment, streaming applications when data is not ready, or visual appearance in
human-readable protocols. Similarly, Nail does not preserve the layout of
objects referred to by their offsets. If the grammar contains no constants and
offset fields, there is a proper isomorphism between model and protocol.

Nail evaluates the combinators and produces:

\begin{itemize}
\item type declarations for the internal model,
\item the \textit{parser}, a function to parse a sequence of bytes into an
instance of the model, and
\item the \textit{generator}, a function to create a
sequence of bytes from an instance of the model.
\end{itemize}

\noindent
In the rest of this section, we present Nail's combinators.

\begin{figure}[tb]
\begin{tabular}{ll}
\toprule
\bf Syntax & \bf Semantics \\
\midrule
\texttt{int32} & 32-bit signed integer \\
\multirow{2}{*}{\texttt{uint4}}
  & 4-bit unsigned integer, \\
  & returned as an 8-bit value \\
\multirow{2}{*}{\texttt{uint8 | 1..3}}
  & 8-bit unsigned integer, \\
  & $1\leq x \leq 3$ \\
\multirow{2}{*}{\texttt{uint16 | ..512}}
  & unsigned 16-bit integer, \\
  & $x \leq 512$ \\
\multirow{2}{*}{\texttt{int32 | [1,255,512]}}
  & signed 32 bit integer, \\
  & $x \in \{ 1, 255, 512 \}$ \\
\bottomrule
\end{tabular}
\caption{Example Nail grammars for integer values.}
\label{fig:range}
\end{figure}

\paragraph{Fundamental parsers.}
The elementary parsers of Nail are the same as those of Hammer, signed
and unsigned integers with arbitrary lengths up to 64 bits.  Note that is
possible to define parsers for sub-byte lengths, e.g. to parse the 4-bit
data offset within the TCP header; in Nail's syntax.  Integer parsers
return their value in the smallest appropriately-sized machine integer
type; e.g., a 24-bit integer is stored in a 32-bit wide variable.

Integer parsers can be constrained to fall either within an (inclusive)
range of values or be an element of a set of allowed values.
Figure~\ref{fig:range} shows several examples.
Invalid values or prematurely reaching the end of input will raise an
error when parsing input or generating output, and abort the procedure.

\paragraph{Sequence.}

Nail's fundamental concept is the structure combinator. It contains a list of
named parsers and unnamed constant parsers. The parser invokes each field 
in sequence and returns a structure containing the result of each parser.
For example, Figure~\ref{fig:dns-struct} shows the structure combinator
from a part of the grammar for DNS packets, along with the data model
corresponding to it.

\begin{figure}[tb]

\begin{minipage}{0.45\columnwidth}
\begin{verbatim}
header = {
  id uint16
  qr uint1
  opcode uint4
  aa uint1 
  tc uint1
  rd uint1
  ra uint1
  uint3 = 0
  rcode uint4
}
\end{verbatim} 
\end{minipage}
~
\begin{minipage}{0.45\columnwidth}
\begin{verbatim}
struct header {
  uint16_t id;
  uint8_t qr;
  uint8_t opcode;
  uint8_t aa;
  uint8_t tc;
  uint8_t rd;
  uint8_t ra;
  uint8_t rcode;
};
\end{verbatim} 
\end{minipage}

\caption{Nail grammar (left) and data model (right) for a part of the
grammar for DNS packets.  The \texttt{uint3 = 0} grammar represents
3 bits of padding (filled with zeroes).}
\label{fig:dns-struct}
\end{figure}

\paragraph{Repetition.}

The \emph{many} combinator takes a parser and applies it repeatedly
until it fails, returning an array of the inner parsers results. The
\emph{sepBy} combinator
additionally takes a constant parser, which it applies in between parsing
two values, but not before parsing the first value or after parsing the
last.
For example, \texttt{many uint8} represents an array of 8-bit unsigned
integers, and \texttt{sepBy uint8=',' (many uint8 | '0'..'9')} recognizes
comma-separated lists of decimal numbers.

\paragraph{Semantic choice.}

We extend a parsing expression grammar's ordered choice combinator with
a tag for each choice. The parser
attempts to parse each option in the order they are specified in the field and
stores the result in a tagged union. If an option fails, the parser backtracks
to the beginning of the choice combinator's input. Care must be taken that the
choices do not overlap, because Nail always picks the first successfully
parsed choice. If two options overlap, generated output for the latter option is
not necessarily understood identically by the parser. However, actual data
formats are usually not ambiguous in this sense.
Figure~\ref{fig:choice} demonstrates a simple choice combinator.

\begin{figure}[tb]
\begin{verbatim}
choose {
  A = uint8 | 1..8
  B = uint16 | ..256
}
\end{verbatim}
\caption{A simple choice combinator that parses either an 8-bit unsigned
integer with a value between 1 and 8 (option A), or a 16-bit unsigned
integer with a value of at most 256 (option B).}
\label{fig:choice}
\end{figure}

\subsection{Constant parsers}

Nail also features \emph{constant parsers}, which are not affecting the internal
representation. Constant parsers can appear instead of structure fields and as
separators in the SepBy combinator.
